Inferring the gravitational potential of the Milky Way with a few precisely measured stars

### 2014-04-08 text/paper2/ms.tex
discussion

@@

For a satellite galaxy orbiting within a static host potential, mass loss is driven by a combination of tidal stripping by the steady tidal field of the parent system and tidal shocking by the rapidly changing tidal field as the progenitor system moves through pericenter \citep[e.g.,][]{choi09}. On a mildly eccentric orbit, the tidal shocking is subdominant to the steady disruption. The interplay between these two processes as a function of orbital properties is outside our scope; we focus here on the slow removal of stars by steady tidal forcing.

@@

We first test the model using only eight stars (particles) --- four from the leading tail, and four from the trailing tail --- randomly sampled from the $2.5\times10^8\,\msun$ simulation (Section~\ref{sec:sims}), assuming the observed 6D positions for both the stars and the progenitor have negligible uncertainties. The stars were required to have been stripped after the first pericentric passage and have a present-day distance within $50$ kpc of the Sun. Figure~\ref{fig:sims} (second column from right) shows the eight randomly chosen stars (black crosses) in Galactic coordinates, over-plotted on all other simulated particles (grey points).

@@

\subsection{Data with near-future uncertainties}\label{sec:exp2}
We next take the same eight stars used in the previous experiment and ``observe'' them with optimistic observational uncertainties. We take these stars to be RR Lyrae variables which are known to be excellent distance indicators via the mid-infrared period-luminosity relation \citep[as shown in, e.g.,][]{madore12}. Nearly 100 RR Lyrae associated with the Sgr stream and $\sim$30 associated with the Orphan stream will be observed with \spitzer as part of the SMASH survey \citep{smashprop} with expected fractional distance uncertainties around $\sim$2\%. These stars will also be included in the \gaia\, proper motion catalog: at a distance of $\sim$50~kpc, a typical RR Lyrae will have a tangential velocity error around $\sim$20~km/s.

For this experiment, we (1) assume the stars are bright distance indicators --- RR Lyrae --- such that the fractional distance uncertainty is $2\%$; (2) neglect the uncertainty in angular position (for a typical RR Lyrae at 50~kpc this is $\sim$$10^{-7}$~deg for \gaia); (3) assume we can measure radial velocities to these stars with 5~km/s uncertainty; and (4) compute the sky-averaged \gaia\, proper-motion uncertainty for each star assuming an F0V spectral type using the \texttt{PyGaia}\footnote{https://github.com/agabrown/PyGaia} code and use this uncertainty for both components of proper motion. We further assume that we know the tail assignment for each star, $\beta$. We observe the position of the progenitor with the same observational uncertainties, though in reality some coordinates will have higher or lower uncertainties.

This experiment samples over the four potential parameters, the Lagrange point offset, four phase-space coordinate parameters plus the unbinding time for each star (40 total), and four phase-space coordinate parameters for the progenitor --- 49 parameters in total. We use an ensemble of 256 walkers and draw initial conditions for the coordinate parameters by sampling from Gaussian's centered on the ``observed'' value with variances specified by the observational uncertainties. Other parameters --- potential parameters, $\Loffset$, and $\tub$ --- are initialized by drawing from the priors summarized in Table~\ref{tbl:params}. We burn in the walkers for 10000 steps and run for 10000 inference steps. Figures~\ref{fig:exp2_potential}, \ref{fig:exp2_satellite}, \ref{fig:exp2_particle} show the marginalized posteriors for the potential parameters, satellite parameters, and parameters for a single particle.

% TODO: slight ~1% bias in values due to particular sample of stars?

@@

\subsection{Current data}\label{sec:exp4}
% Most of the streams and debris structures discovered thus far have undetected or unconfirmed progenitors. We expect the phase-space positions of the stars to provide constraints on the position of the progenitor even if it is unobserved. For this experiment, we consider a situation where we have precise (2\%) distance measurements, 10~km/s radial velocity uncertainties, and no proper motion measurements to eight stream stars (the same used in previous experiments), and no knowledge of the 6D position of the progenitor object. 
Without measured proper motions, the extremely precise distance measurements combined with moderate precision radial velocities are sufficient for constraining the properties of the debris and the potential. How important are these distance measurements? We now test \rewinder\ on just eight stars assuming the best uncertainties available for existing data of the Sgr stream. For the stream stars, we adopt 15\% distance uncertainties \citep[typical for M-giants, e.g.,][]{majewski04}, 0.1~mas/yr proper motion uncertainties \citep{koposov13}, and 5~km/s radial velocity uncertainties \citep{majewski04}. Note that we neglect any systematic uncertainties or trends with metallicity which are known to effect M-giant distance measurement accuracy. For the progenitor (Sgr core), we adopt a 1~kpc distance uncertainty \citep{kunder09}, 0.2~mas/yr proper motion uncertainties \citep{pryor10}, and 10~km/s radial velocity uncertainty \citep{frinchaboy12}. We again neglect uncertainties in the sky coordinates.

Again, this experiment includes 49 model parameters in total. We use an ensemble of 256 walkers and also initialize the walkers by sampling the coordinate parameters from Gaussians centered on the true positions with variances set to 1\% of the parameter values. The potential parameters, $\Loffset$, and $\tub$ are initialized by drawing from the priors summarized in Table~\ref{tbl:params}. We again burn in the walkers for 10000 steps and run for 10000 inference steps. Figures~\ref{fig:exp4_potential}, \ref{fig:exp4_satellite}, \ref{fig:exp4_particle} show the marginalized posteriors for the potential parameters, satellite parameters, and parameters for the same star shown in previous figures. %[TODO: ..statement about results]

@@

\section{Discussion}\label{sec:discussion}
% In Section~\ref{sec:intro} we identified several issues that have only been partially solved by other stream modeling methods; \rewinder\ addresses each of these.
%\begin{enumerate}
%	\item Observational uncertainties and missing data dimensions are naturally handled in the inferences by including the true 6D positions of stars and the progenitor as parameters in the model.
%	\item The orbits of the stars and progenitor are directly integrated and only require a function that evaluates the acceleration due to the potential at a given position (and time). 
%	\item Incorporating multiple streams into the inference only requires multiplying likelihoods computed for each stream and progenitor.
%	\item Each star adds constraints on the potential parameters and does not require matching a simulated density to the observed density of stars.
%	\item The inference is much faster than full N-body modeling because the orbits of the stars and progenitor are integrated as test particles.
%\end{enumerate}
%\begin{enumerate}
%	\item The known debris structures are $\gtrsim$10 kpc from the Sun, thus while overdensities in the halo are less contaminated by background, the constituent stars often have incomplete or poorly measured phase-space information. Thus it is critical for any approximate model of tidal debris to incorporate observational uncertainties and missing dimensions in a consistent and justified way. 
%	\item The potential of the inner galaxy is conceivably well-described by a (set of) nearly-integrable potentials, thus the dynamics of collisionless systems is plausibly captured by a transformation to some integral-of-motion-space, e.g., action-angle coordinates. However there is large uncertainty in the radial profile, shape, orientation, and graininess of the outer halo and the constancy of these parameters over distance. Properly describing the potential may require complex models where computing actions is non-trivial or even impossible.
%	\item Near-future photometric surveys such as \gaia\, and the \project{LSST} will likely discover many new streams and kinematic associations of stars. Potential recovery methods should be able to simultaneously model multiple streams and incorporate other dynamical constraints.
%	\item Matching generated models to observed stream densities is difficult: models that rely on this must account for observational biases and properly model the underlying background of halo stars. 
%	\item Full N-body simulations are expensive to run. Incorporating an N-body model into a general model and then performing a parameter search would be computationally intensive and, presently, intractable. 
%\end{enumerate}
% are naturally handled in the inferences 
Any method that uses tidal debris as a potential measure must also model the uncertainties on the kinematic measurements because the observational uncertainties will be significant. \rewinder\ handles observational uncertainties and missing data dimensions by including the true 6D positions of stars and the progenitor as parameters in the model. In this article we have tested \rewinder\ using data of varied quality for a small sample of stars. In future work, we will investigate the effect of reducing the data quality but increasing the sample size.

No assumptions were made about the the form of the Galactic potential used in \rewinder: the potential could be time-dependent, clumpy, or have properties that vary with radius. The orbits of the stars and progenitor are directly integrated and only require a function that evaluates the acceleration due to the potential at a given position (and time). The simple experiments in this article infer potential parameters from the same functional potential used in the N-body simulations that generated the fake data. In future work, we will (1) try fitting incorrect potential models to see if we can still recover global properties, e.g., flattening; (2) run simulations with smoothly changing potentials \citep[e.g.,][]{buist14} and attempt to model the time dependence; and (3) explore using a generic, non-parametric potential form (e.g., a basis function expansion) as the recovery potential model.

Each stream will provide constraints on different properties of the potential. For example, more eccentric streams may better constrain the radial profile \citep[see][who illustrate the power of using multiple streams to simultaneously constrain the potential using orbit fitting]{deg14}. In this article, we only consider a stream on a Sgr-like orbit, which lies nearly in the Galactic $x$-$z$ plane. We find that the stream puts better constraints on the $z$-axis flattening, $q_z$, than on the flattening in the $x$-$y$ plane, $q_1$, as expected. The best measurements of the detailed shape of the Galactic potential will come from combining constraints from multiple streams. We have defined above a proper likelihood function for observed stars in a given stream, thus incorporating multiple streams into the inference only requires multiplying the likelihoods computed for each stream and progenitor pair.

The inference is much faster than full N-body modeling because the stars and progenitor are treated as test particles. Each step in parameter space with \rewinder\ requires integrating the orbits of stars and the progenitor: we presently integrate with a fixed timestep using leapfrog integration. Computing the acceleration due to the given potential at each step is implemented in \project{Cython} (and approaches \project{C}-like speed), but the rest of the code is written in pure-\project{Python}. Though already significantly less computationally intensive than running a full N-body simulation for every parameter step, running each MCMC walker for 10000 steps requires parallelization and a few hours of CPU time on a compute cluster. Further optimization of the integration method (e.g., using an adaptive method or implementing in \project{Cython}) could speed up the inference significantly.

One point of concern is that the number of parameters scales with eight times the number of stars: each star has six coordinate parameters (the true position), the unbinding time, and the tail assignment. Computational constraints limit the sample size, although it might be that the most relevant stellar samples for inferring the Milky Way potential are small. For instance, stars that produce good distance estimates might be much more valuable than typical stars in the sample so that we may limit to only variable stars, e.g., RR Lyraes. These valuable stars are rare (and their abundances are age and metallicity dependent); there could be many cold structures in the Milky Way halo that are highly constraining on the potential in principle, but which contain only a few good distance-indicating members. With \rewinder, each star adds constraints on the potential parameters and does not require matching a simulated density to the observed density of stars, thus this method works well even for small samples of well-measured stream stars. That said, it is not yet known what the trade-offs are between having many stars at low precision and a few at high precision, nor is it known how valuable distance information really is, when a structure contains many precisely observed members.

Another reason to think about small samples is more conceptual:
in dynamical inferences that go beyond simple virial or Jeans arguments,
  inference depends on the non-linear dynamical evolution of populations in phase space.
The nonlinearities, 
  combined with heteroskedastic data and requirements to marginalize over phase-space distribution functions,
  lead naturally to situations in which some stars are \emph{far more informative} than others.
For a simple example, in a toy study of the Solar System,
  phase-space distribution marginalization was shown to make some planets are far more informative than others
  about the gravitational potential \citep{bovy10}.
This is expected to be a generic feature in dynamical problems:
  it might be that a very small fraction of \gaia's catalog entries
  deliver a large fraction of the information about the Milky Way's potential and formation history.

In this article, we have only applied \rewinder\ to simulated observations of a kinematically ``hot'' stream; a large mass satellite will have a large dispersion in both position and velocity space. It has been shown previously that simulated cold streams are better than hot streams at recovering potential properties (cite?), however it is not clear whether this remains true with realistic observational uncertainties on the stream stars. For instance, when the uncertainties ``blur'' out the kinematic overdensity that is the stream, it is not known whether a cold stream would still be more informative than an equal sample of stars in a hot stream such as Sgr. This blurring could also make it more difficult to determine stream membership.

This model works well when all stars are lost due to tidal pruning. Even on a mildly eccentric orbit, stars are preferentially stripped around the satellite's pericenter because of tidal shocking. We will explore properly modeling this process in future work. We have focused on modeling streams of stars formed from a disrupting satellite on a mildly eccentric orbit. On more eccentric orbits, the debris tends to form more shell or cloud-like structures. The morphological distinction between stellar streams and stellar clouds is just related to the orbit of the disrupting satellite within the parent galaxy. With a proper model for the tidal shocking that occurs on an eccentric orbit, \rewinder\ could also be applied to structures such as the Triangulum-Andromeda (cite) or Hercules-Aquila (cite) stellar clouds, both possible shell-like structures viewed from the inside.

>>>>

\rewinder\ integrates stars and the progenitor back from their present-day, observed positions to the time at which they become unbound from the satellite, where we evaluate the likelihood for each star. Below we (1) motivate a simple model for the debris just as it disrupts (near $\tub$ for each star) and (2) present this idea within a probabilistic framework for incorporating observational uncertainties and missing data. For a satellite galaxy orbiting within a static host potential, mass loss is driven by a combination of tidal stripping by the steady tidal field of the parent system and tidal shocking by the rapidly changing tidal field as the progenitor system moves through pericenter \citep[e.g.,][]{choi09}. On a mildly eccentric orbit, the tidal shocking is subdominant to the steady disruption. The interplay between these two processes as a function of orbital properties is outside our scope; we focus here on the slow removal of stars by steady tidal forcing.

@@

@@

We first test \rewinder\ using only eight stars (particles) --- four from the leading tail, and four from the trailing tail --- randomly sampled from the $2.5\times10^8\,\msun$ simulation (Section~\ref{sec:sims}), assuming the observed 6D positions for both the stars and the progenitor have negligible uncertainties. The stars were required to have been stripped after the first pericentric passage and have a present-day distance within $50$ kpc of the Sun. Figure~\ref{fig:sims} (second column from right) shows the eight randomly chosen stars (black crosses) in Galactic coordinates, over-plotted on all other simulated particles (grey points).

@@

\subsection{Data with near-future uncertainties}\label{sec:exp2}
We next take the same eight stars used in the previous experiment and ``observe'' them with optimistic observational uncertainties. We require these stars to be RR Lyrae variables which are known to be excellent distance indicators via the mid-infrared period-luminosity relation \citep[as shown in, e.g.,][]{madore12}. Nearly 100 RR Lyrae associated with the Sgr stream and $\sim$30 associated with the Orphan stream will be observed with \spitzer as part of the SMASH survey \citep{smashprop} with expected fractional distance uncertainties around $\sim$2\%. These stars will also be included in the \gaia\, proper motion catalog: at a distance of $\sim$50~kpc, a typical RR Lyrae will have a tangential velocity error around $\sim$20~km/s.

For this experiment, we (1) assume the stars are bright distance indicators --- RR Lyrae --- such that the fractional distance uncertainty is $2\%$; (2) neglect the uncertainty in angular position (for a typical RR Lyrae at 50~kpc this is $\sim$$10^{-7}$~deg for \gaia); (3) assume we can measure radial velocities to these stars with 5~km/s uncertainty; and (4) compute the sky-averaged \gaia\, proper-motion uncertainty for each star assuming an F0V spectral type using the \texttt{PyGaia}\footnote{https://github.com/agabrown/PyGaia} code and use this uncertainty for both components of proper motion. We further assume that we know the tail assignment for each star, $\beta$. We observe the position of the progenitor with the same observational uncertainties though in reality some coordinates will have even better measurements.

This experiment samples over the four potential parameters, the Lagrange point offset, four phase-space coordinate parameters plus the unbinding time for each star (40 total), and four phase-space coordinate parameters for the progenitor --- 49 parameters in total. We use an ensemble of 256 walkers and draw initial conditions for the coordinate parameters by sampling from Gaussian's centered on the ``observed'' value with variances specified by the observational uncertainties. Other parameters --- potential parameters, $\Loffset$, and $\tub$ --- are initialized by drawing from the priors summarized in Table~\ref{tbl:params}. [TODO: start from truth...] We burn in the walkers for 10000 steps and run for 10000 inference steps. Figures~\ref{fig:exp2_potential}, \ref{fig:exp2_satellite}, \ref{fig:exp2_particle} show the marginalized posteriors for the potential parameters, satellite parameters, and parameters for a single particle.

% TODO: slight bias in values due to particular sample of stars?

@@

\subsection{Current data}\label{sec:exp4}
% Most of the streams and debris structures discovered thus far have undetected or unconfirmed progenitors. We expect the phase-space positions of the stars to provide constraints on the position of the progenitor even if it is unobserved. For this experiment, we consider a situation where we have precise (2\%) distance measurements, 10~km/s radial velocity uncertainties, and no proper motion measurements to eight stream stars (the same used in previous experiments), and no knowledge of the 6D position of the progenitor object. 
Without measured proper motions, the extremely precise distance measurements combined with moderate precision radial velocities are sufficient for constraining the properties of the debris and the potential. How important are these distance measurements? We now test \rewinder\ on just eight stars assuming the best uncertainties available for existing data of the Sgr stream. For the stream stars, we adopt 15\% distance uncertainties \citep[typical for M-giants, e.g.,][]{majewski04}, 0.1~mas/yr proper motion uncertainties \citep{koposov13}, and 5~km/s radial velocity uncertainties \citep{majewski04}. Note that we neglect any systematic uncertainties or trends with metallicity which are known to effect M-giant distance measurement accuracy. For the progenitor (Sgr core), we adopt a 1~kpc distance uncertainty \citep{kunder09}, 0.2~mas/yr proper motion uncertainties \citep{pryor10}, and 10~km/s radial velocity uncertainty \citep{frinchaboy12}. We  neglect uncertainties in the sky coordinates.

Again, this experiment includes 49 model parameters in total. We use an ensemble of 256 walkers and also initialize the walkers by sampling the coordinate parameters from Gaussians centered on the true positions with variances set to 1\% of the parameter values. The potential parameters, $\Loffset$, and $\tub$ [TODO: truth] are initialized by drawing from the priors summarized in Table~\ref{tbl:params}. We again burn in the walkers for 10000 steps and run for 10000 inference steps. Figures~\ref{fig:exp4_potential}, \ref{fig:exp4_satellite}, \ref{fig:exp4_particle} show the marginalized posteriors for the potential parameters, satellite parameters, and parameters for the same star shown in previous figures. %[TODO: ..statement about results]

@@

\section{Discussion}\label{sec:discussion}
{\bf 1. Observational uncertainties:} Any method that uses tidal debris as a potential measure must also model the (significant) uncertainties on the kinematic measurements; \rewinder\ handles observational uncertainties and missing data dimensions by including the true 6D positions of stars and the progenitor as model parameters. In this article we have tested \rewinder\ using data of varied quality for a small sample of stars.

{\bf 2. Form of the potential:} No assumptions were made about the the form of the Galactic potential used in \rewinder: the potential could be time-dependent, clumpy, or have properties that vary with radius. The orbits of the stars and progenitor are directly integrated and only require a function that evaluates the acceleration due to the potential at a given position (and time). The simple experiments in this article infer potential parameters from the same functional potential used in the N-body simulations that generated the fake data. In future work, we will (1) try fitting incorrect potential models to see if we can still recover global properties (e.g., flattening); (2) run simulations with smoothly changing potentials \citep[e.g.,][]{buist14} and attempt to model the time dependence; and (3) explore using a generic, non-parametric potential form (e.g., a basis function expansion) as the recovery potential model.

{\bf 3. Multiple debris structures:} Each stream will provide constraints on different properties of the potential. For example, more eccentric streams may better constrain the radial profile \citep[see][who illustrate the power of using multiple streams to simultaneously constrain the potential using orbit fitting]{deg14}. In this article, we only consider a stream on a Sgr-like orbit, which lies nearly in the Galactic $x$-$z$ plane. We find that the stream puts better constraints on the $z$-axis flattening, $q_z$, than on the flattening in the $x$-$y$ plane, $q_1$, as expected. The best measurements of the detailed shape of the Galactic potential will come from combining constraints from multiple streams. We have defined above a proper likelihood function for observed stars in a given stream, thus incorporating multiple streams into the inference only requires multiplying the likelihoods computed for each stream and progenitor pair.

{\bf 4. Comparing models to data:} With \rewinder, each star adds constraints on the potential parameters and does not require matching a simulated density to the observed density of stars, thus this method works well even for small samples of well-measured stream stars. It might be that the most relevant stellar samples for inferring the Milky Way potential are small. For instance, stars that produce good distance estimates might be much more valuable than typical stars in the sample so that we may limit to only variable stars, e.g., RR Lyraes. These valuable stars are rare (and their abundances are age and metallicity dependent); there could be many cold structures in the Milky Way halo that are highly constraining on the potential in principle, but which contain only a few good distance-indicating members. It is not yet known what the trade-offs are between having many stars at low precision and a few at high precision, nor is it known how valuable distance information really is, when a structure contains many precisely observed members.

{\bf 5. Computational expense:} One point of concern is that the number of parameters scales with eight times the number of stars: each star has six coordinate parameters (the true position), the unbinding time, and the tail assignment. Computational constraints limit the sample size, but even so, the inference is much faster than full N-body modeling because the stars and progenitor are treated as test particles. Each step in parameter space with \rewinder\ requires integrating the orbits of stars and the progenitor: we presently integrate with a fixed timestep using leapfrog integration. Computing the acceleration due to the given potential at each step is implemented in \project{Cython} (and approaches \project{C}-like speed), but the rest of the code is written in pure-\project{Python}. Though already significantly less computationally intensive than running a full N-body simulation for every parameter step, running each MCMC walker for 10000 steps requires parallelization and a few hours of CPU time on a compute cluster. Further optimization of the integration method (e.g., using an adaptive method or implementing in \project{Cython}) could speed up the inference significantly.

%Another reason to think about small samples is more conceptual:
%in dynamical inferences that go beyond simple virial or Jeans arguments,
%  inference depends on the non-linear dynamical evolution of populations in phase space.
%The nonlinearities, 
%  combined with heteroskedastic data and requirements to marginalize over phase-space distribution functions,
%  lead naturally to situations in which some stars are \emph{far more informative} than others.
%For a simple example, in a toy study of the Solar System,
%  phase-space distribution marginalization was shown to make some planets are far more informative than others
%  about the gravitational potential \citep{bovy10}.
%This is expected to be a generic feature in dynamical problems:
%  it might be that a very small fraction of \gaia's catalog entries
%  deliver a large fraction of the information about the Milky Way's potential and formation history.

%In this article, we have only applied \rewinder\ to simulated observations of a kinematically ``hot'' stream; a large mass satellite will have a large dispersion in both position and velocity space. It has been shown previously that simulated cold streams are better than hot streams at recovering potential properties (cite?), however it is not clear whether this remains true with realistic observational uncertainties on the stream stars. For instance, when the uncertainties ``blur'' out the kinematic overdensity that is the stream, it is not known whether a cold stream would still be more informative than an equal sample of stars in a hot stream such as Sgr. This blurring could also make it more difficult to determine stream membership.

\rewinder works well when all stars are lost due to tidal pruning. Even on a mildly eccentric orbit, stars are preferentially stripped around the satellite's pericenter because of tidal shocking. We will explore properly modeling this process in future work. We have focused on modeling streams of stars formed from a disrupting satellite on a mildly eccentric orbit. On more eccentric orbits, the debris tends to form more shell or cloud-like structures. The morphological distinction between stellar streams and stellar clouds is just related to the orbit of the disrupting satellite within the parent galaxy. With a proper model for the tidal shocking that occurs on an eccentric orbit, \rewinder\ could also be applied to structures such as the Triangulum-Andromeda (cite) or Hercules-Aquila (cite) stellar clouds, both possible shell-like structures viewed from the inside.
